Sampling Strategies in Decisions from Experience

### 2022-02-15 manuscript.Rmd
Update: Review of structure

The human mind is a cognitive system that operates on inputs from an environment or memory.
In the domain of decision making, such inputs provide information about the choice alternatives among which people can choose.
Let us treat these choice alternatives hereafter as elements $c$ of the set $C$.
To each $c \in C$ is then assigned some information from the inputs and this information may be used by the mind to learn about the *properties* of the respective choice alternative.
In turn, decision making can be considered an information processing problem which starts from the set $C$, containing all choice alternatives, and stops at a proper subset $C' \subset C$ with the inclusion map
$$
\begin{aligned}
  \iota : C' &\to C \\
  c &\mapsto c  \quad \forall c \in C'
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, p. 15], containing only the chosen alternatives.
Indeed, irrespective of their exact mathematical specification or computational implementation, decision models generally take as inputs information about a set of choice alternatives to return a subset thereof as outputs [@heOntologyDecisionModels2020].

<!-- Add short and general outlook on the paper -->

@@

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, Chapter 1, for a timeless discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered the possible outcomes of a choice alternative---hereafter $x_i$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p_X(x_i)$.
Roughly speaking, then, all choice alternatives that can be fully described by their outcome-probability pairs $(x_i, p_X(x_i))$ are prospects (or: gambles), which are exclusively considered below. 
<!--For illustration, take the choice alternatives in the following example with monetary outcomes:

@@

Now, more formally, let the axiomatic definition of a prospect be that of a discrete random variable's distribution.
Let therefore the tuple $(\Omega, \mathcal{F}, P)$ be a probability space. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ is the sample space containing a finite set of elements
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega) = \mathcal{F}$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \to [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.<!--^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \to [0,1]$, which is used to construct the probability measures $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]-->
The random variable $X$ is the measurable function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{4}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ being a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the probability measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{5}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(X^{-1} \ A') = P(\{\omega: X(\omega) = x \in A'\})
\tag{6}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21; see also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1, for a comprehensive introduction to random variables].
That is, the set consisting of a single outcome $x$ can be treated as a subset $\{x\} \in \mathcal{F'}$ and we therefore obtain $p_X(x) = P(X(\omega) = x)$.
Note that the probability measure in (5) and (6), respectively, is a discrete random variable's distribution and therefore a prospect.
To summarize and simplify, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
C = \{P_{X_1}, ... , P_{X_k}\}
\tag{7}
$$
with each element $c \in C$ being a distribution $P_X$ of a random variable defined on a different probability space.
Accordingly, each $c$ can itself be considered a set of outcome-probability pairs $(x_i, p_X(x_i)) \in \{(x_1, p_X(x_1)), ..., (x_n, p_X(x_n))\}$.

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes-probability pairs.
In such cases, people would make *decisions from description* (DfD), since the information from which the mind learns about the prospects are complete descriptions thereof.
Yet, rather often, people make *decisions from experience* [DfE, @hertwigDecisionsExperienceEffect2004], where the mind can learn about the prospects only from a record of past experiences with them.
That is, people may have experienced the possible outcomes of a prospect when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular prospect is a rough description of a *sample*.
The mind may then infer from the relative frequency with which an outcome occurred in such a sample of past decisions its probability of occurrence following a future decision.

@@

The choices people actually make between prospects---be that in DfD or DfE---are often described in terms of deviations from expected value (EV) maximization, where the EV of a prospect is given by 
$$
EV = \sum_i^n x_i \times p_X(x_i)
\; .
\tag{9}
$$ 
That is, following the principle of EV maximization, one would calculate the prospects' EV---each by first multiplying its outcomes $x_i$ by their respective probabilities $p_X(x_i)$ and then summing the resulting products---to subsequently choose the prospect with the largest EV.
To *describe* how people's actual choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing the value
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{10}
$$ 
with objective outcomes $x_i$ being transformed by a *value function* $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P_X$.^[Note that CPT or other originally deterministic decision models may be amended by a stochastic choice function to account for the probabilistic nature of choices [see, e.g., @rieskampProbabilisticNaturePreferential2008].]
Importantly, the transformation of cumulative probabilities is carried out by a weighting function $w$ (see (11) and (12) below).
More specifically, CPT proposes cumulative decision weights as replacements for objective probabilities to account for empirical regularities in the deviations from EV maximization in DfD^[@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects such as the *certainty effect* and the *reflection effect* that were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the EV principle and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].] that cannot be captured by the transformation of objective outcomes into subjective values $v(x)$ alone (see Footnote 6).
In essence, to provide a proper description of people's choices between prospects in terms of their deviation from EV maximization, the proposed model is such that the inferred cumulative decision weights may be larger or smaller than the respective objective probabilities in (9); and they may therefore give more or less weight to subjective values for the overall valuation of a prospect in CPT than would be objectively warranted.
In other words, for an inferred cumulative decision weight that is greater (smaller) than the objective probability it replaces, the respective outcome is overweighted (underweighted) in CPT.
Accordingly one may---but must not---adopt an as-if weighting terminology for the description of choices by stating that people decide *as if* they maximized the value in (10) and applied the inferred over- and underweighting pattern.
Importantly, the as-if prefix indicates that it cannot be concluded from the inferred cumulative decision weights that the mind does indeed perform the weighting computation in its literal sense, but it processes the inputs in a way that the outputs approximate the solutions of what can be considered probability weighting on an abstract computational level [see, e.g., @griffithsRationalUseCognitive2015].

>>>>

The human mind is a cognitive system that operates on inputs from an environment or memory.
In the domain of decision making, such inputs provide information about the *properties* of the choice alternatives among which people can choose.
Hereafter, let us treat these alternatives as elements $c$ of the set $C$.
To each $c \in C$ is then assigned some information from the inputs and this information may be used by the mind to learn about the properties of the respective alternative.
Accordingly, decision making can be considered an information processing problem which starts from the set $C$, containing all choice alternatives, and stops at a proper subset $C' \subset C$ with the inclusion map
$$
\begin{aligned}
  \iota : C' &\to C \\
  c &\mapsto c  \quad \forall c \in C'
\end{aligned}
\tag{1}
$$
[@jostMathematicalConcepts2015, p. 15], containing only the chosen alternatives.
Indeed, irrespective of their exact mathematical specification or computational implementation, decision models generally take as inputs information about a set of choice alternatives to return a subset thereof as outputs [@heOntologyDecisionModels2020].

Importantly, information about the choice alternatives can come in many formats.
In *decisions from description* (DfD), inputs take the form of explicit and complete descriptions of all relevant properties of the choice alternatives.
In *decisions from experience* [DfE, @hertwigDecisionsExperienceEffect2004], inputs take the form of *samples* (see below).
In this latter case---perhaps, the one that more closely resembles decisions in people's daily life---the properties of the choice alternatives are not described and therefore not known with certainty, but they must be inferred from the consequences that occurred or were absent after the alternatives were chosen in similar decisions in the past.
Although these different formats may in principle carry the same information about given choice alternatives, behavioral decision research has yet produced a large body of papers indicating that, taking a quite general perspective, there are robust differences between the choices that are made in DfD and DfE.
That is, all other things being equal, empirical choice patterns across studies indicate that in DfD people choose *as if* a rare consequence of a choice alternative has more impact on the latter's evaluation than would be objectively warranted by the principle of mathematical expectation (see below); in turn, in DfE, people choose *as if* the same rare consequence has less impact on the evaluation than warranted [e.g., @barronSmallFeedbackbasedDecisions2003; @hertwigDecisionsExperienceEffect2004; @weberPredictingRiskSensitivity2004; @hauDescriptionexperienceGapRisky2008; @rakowBiasedSamplesNot2008; @ungemachAreProbabilitiesOverweighted2009; @camilleriWhenWhyRare2011; @kellenHowVariantAre2016; @erevAnomaliesForecastsDescriptive2017; @regenwetterConstructBehaviorGap2017; see also @wulffMetaanalyticReviewTwo2018, for a comprehensive meta-analytic review].

This paper uses the *description-experience gap* [@hertwigDescriptionexperienceGapRisky2009] as an entry point for a closer inquiry into the sampling process as an essential part of sample-based decision making.

<!--
The paper is structured as follows: 
...

All simulations and analyses are reproducible---materials and instructions can be found on the GitHub repository.
-->

@@

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from decisions between particular choice alternatives, e.g., the choice between job offers, political parties to vote for, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (see below), it studies a case which is counterfactual in that it omits the many particularities of each decision, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See @knightRiskUncertaintyProfit1921, Chapter 1, for a discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered the possible outcomes of a choice alternative---hereafter $x_i$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p_X(x_i)$.
Roughly speaking, then, all choice alternatives that can be fully described by their outcome-probability pairs $(x_i, p_X(x_i))$ are prospects (or: gambles), which are exclusively considered below. 
<!--For illustration, take the choice alternatives in the following example with monetary outcomes:

@@

Now, more formally,^[Paragraphs with this beginning usually formalize the main ideas of the preceding paragraphs or elaborate on them by using mathematical expressions and concepts. Although important, the gist of this paper should be captured despite skipping these paragraphs.]  let the axiomatic definition of a prospect be that of a discrete random variable's distribution.
Let therefore the tuple $(\Omega, \mathcal{F}, P)$ be a probability space. 
Following the @kolmogorovFoundationsTheoryProbability1950 axioms, $\Omega$ is the sample space containing a finite set of elements
$$
\omega_i \in \{\omega_1, ..., \omega_n\} = \Omega
\; .
\tag{2}
$$ 
$\mathcal{F}$ is the set of all possible subsets of $\Omega$, i.e., the power set $\mathcal{P}(\Omega)$ with each element $A \in P(\Omega) = \mathcal{F}$ being a subset of $\Omega$.^[Note that this paper considers only the case of finite sample spaces. The exact axiomatic definition of $\mathcal{F}$, which can be found in Kolmogorov [-@kolmogorovFoundationsTheoryProbability1950, pp. 2-3 and 14-15], also covers infinite sample spaces and is that of a $\sigma$-algebra. For the finite case, however, $\mathcal{P}(\Omega)$ can be considered as such [@georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1.1].] 
$P$ is a probability measure
$$
P: \mathcal{F} \to [0,1]
\; ,
\tag{3}
$$
which assigns each subset $A \in \mathcal{F}$ a probability $0 < p(A) \leq 1$ with $P(\Omega) = 1$.<!--^[The probabilities $p(\omega)$ are determined by the probability mass function $p : \Omega \to [0,1]$, which is used to construct the probability measures $P(A) = \sum_{\omega \in A} p(\omega)$ for all $A \in \mathcal{F}$.]-->
The random variable $X$ is the measurable function
$$
\begin{aligned}
  X : (\Omega, \mathcal{F}) &\to (\Omega', \mathcal{F'})  \\
  \omega &\mapsto x
  \; ,
\end{aligned}
\tag{4}
$$
where $(\Omega', \mathcal{F'})$ is a measurable space with each element $x \in \Omega'$ being a possible outcome of a prospect and each element $A' \in \mathcal{F'}$ being a subset of $\Omega'$.
An outcome's probability $p_X(x)$ is then provided by the probability measure
$$
P_X : \mathcal{F'} \to [0,1]
\tag{5}
$$
with 
$$
P_X(A' \in \mathcal{F'}) := P(X^{-1} \ A') = P(\{\omega: X(\omega) = x \in A'\})
\tag{6}
$$
[see @kolmogorovFoundationsTheoryProbability1950, p. 21; see also @georgiiStochastikEinfuhrungWahrscheinlichkeitstheorie2015, Chapter 1, for a comprehensive introduction to random variables].
That is, the set consisting of a single outcome $x$ can be treated as a subset $\{x\} \in \mathcal{F'}$ and we therefore obtain $p_X(x) = P(X(\omega) = x)$.
Note that the probability measure in (5) and (6), respectively, is a discrete random variable's distribution and therefore a prospect.
To summarize and simplify, given a choice between $1, ..., k$ prospects, the inputs from the environment or memory should provide information about the set
$$
C = \{P_{X_1}, ... , P_{X_k}\}
\tag{7}
$$
with each element $c \in C$ being a distribution $P_X$ of a random variable defined on a different probability space.
Accordingly, each $c$ can itself be considered a set of outcome-probability pairs $(x_i, p_X(x_i)) \in \{(x_1, p_X(x_1)), ..., (x_n, p_X(x_n))\}$.

However, rarely in peoples' daily life, inputs take the form of explicit descriptions of all outcomes-probability pairs.
In such cases, people would make *decisions from description* (DfD), since the information from which the mind learns about the prospects are complete descriptions thereof.
Yet, rather often, people make *decisions from experience* [DfE, @hertwigDecisionsExperienceEffect2004], where the mind can learn about the prospects only from a record of past experiences with them.
That is, people may have experienced the possible outcomes of a prospect when they themselves or others have made the same decision repeatedly in the past, and the record of experienced outcomes for this particular prospect is a rough description of a sample.
The mind may then infer from the relative frequency with which an outcome occurred in such a sample of past decisions its probability of occurrence following a future decision.

@@

The choices people actually make between prospects---be that in DfD or DfE---are often described in terms of deviations from expected value (EV) maximization, where the EV of a prospect is given by 
$$
EV = \sum_i^n x_i \times p_X(x_i)
\; .
\tag{9}
$$ 
That is, following the principle of EV maximization, one would calculate the prospects' EV---each by first multiplying its outcomes $x_i$ by their respective probabilities $p_X(x_i)$ and then summing the resulting products---to subsequently choose the prospect with the largest EV.
To *describe* how people's actual choices deviate from EV maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; @wakkerAxiomatizationCumulativeProspect1993] and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing the value
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{10}
$$ 
with objective outcomes $x_i$ being transformed by a *value function* $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P_X$.^[Note that CPT or other originally deterministic decision models may be amended by a stochastic choice function to account for the probabilistic nature of choices [see, e.g., @rieskampProbabilisticNaturePreferential2008].]
Importantly, the transformation of cumulative probabilities is carried out by a weighting function $w$ (see (11) and (12) below).
More specifically, CPT proposes cumulative decision weights as replacements for objective probabilities to account for empirical regularities in the deviations from EV maximization in DfD^[@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects such as the *certainty effect* and the *reflection effect* that were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the EV principle and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].] that cannot be captured by the transformation of objective outcomes into subjective values $v(x)$ alone (see Footnote 6).
In essence, to provide a proper description of people's choices between prospects in terms of their deviation from EV maximization, the proposed model is such that the inferred cumulative decision weights may be larger or smaller than the respective objective probabilities in (9); and they may therefore give more or less weight to subjective values for the overall valuation of a prospect in CPT than would be objectively warranted.
In other words, for an inferred cumulative decision weight that is greater (smaller) than the objective probability it replaces, the respective outcome is overweighted (underweighted) in CPT.
Accordingly one may adopt an as-if weighting terminology for the description of choices by stating that people decide *as if* they maximized the value in (10) and applied the inferred over- and underweighting pattern [@pachurWeightUncertainEvents2019].
Importantly, the as-if prefix indicates that it cannot be concluded from the inferred cumulative decision weights that the mind does indeed perform any of the computations associated with the weighting of outcomes in CPT; rather, it processes the inputs in a way that the outputs, i.e., the resulting choice patterns, approximate the solutions from the inferred CPT model [see, e.g., @griffithsRationalUseCognitive2015]. 
Below, CPT will therefore only be used as a model to describe choices between prospects and it will be discussed---and demonstrated in a simulation---how distinct strategies of generating and integrating samples in DfE can contribute to weighting patterns that are commonly inferred when modeling respective empirical choice data.
